Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)\\ Technical Appendix Including Proofs and Details

### 2017-02-22 local-capability-machine.tex
added malloc and crtcls macro correctness lemmas.

% Conv defs
\newcommand{\lookingat}[3]{\ensuremath{#1} \text{ is looking at } \ensuremath{#2} \text{ followed by } \ensuremath{#3}}
\newcommand{\pointstostack}[3]{\ensuremath{#1} \text{ points to stack with } \ensuremath{#2} \text{ used and } \ensuremath{#3} \text{ unused}}
\newcommand{\nonlocal}[1]{\ensuremath{#1} \text{ is non-local}}

@@

@@

              \subsection{Macro definitions}
              \begin{description}
              \item[\texttt{fetch} $r$ $f$] load the entry of the linking table corresponding to $f$ to register $r$.
              \item[\texttt{call} $r(\bar{r}_{\var{args}},\bar{r}_{\var{priv}})$] \forcenewline
                $\bar{r}_{\var{args}}$ and $\bar{r}_{\var{priv}}$ are lists of registers. An overview of this call:
                \begin{itemize}
                \item Set up activation record
                \item Create local enter capability for activation (protected return pointer)
                \item Clear unused registers
                \item Jump
                \item Upon return: Run activation code
                \end{itemize}
                A more detailed description of each of the above steps:
                \begin{description}
                \item [Set up activation record]\forcenewline
                  \begin{itemize}
                  \item Run malloc to get a piece of memory with space for:
                    \begin{itemize}
                    \item Words in $\bar{r}_{\var{priv}}$
                    \item Code return capability (opc)
                    \item Activation code
                    \end{itemize}
                  \item Store the words in $\bar{r}_{\var{priv}}$ to the activation record. 
                  \item Adjust a copy of the current pc to point to the return address in code and save it to the activation record.
                  \item Write the activation code to the activation record.
                  \end{itemize}
                \item [Create local enter capability for activation] Adjust the capability for the activation record to point to the beginning of the activation record and restrict it to a local enter-capability. Place this capability in $r_0$.
                \item [Clear unused registers]
                  Clear all the register that are not $\pcreg$, $r$, $r_0$ or in $\bar{r}_{\var{args}}$.
                \item [Jump] Jump to register $r$
                \item [Activation code] The activation code does the following:
                  \begin{itemize}
                  \item Move the stored ``private'' words in to their respective $\bar{r}_{\var{priv}}$ registers.
                    % \item Clear all registers but $r_1$, $\pcreg$, and $\bar{r}_{\var{priv}}$.
                  \item Load the return capability to $\pcreg$
                  \end{itemize}
                \end{description}
              \item[\texttt{malloc $r$ $n$}] Calls malloc to allocates a piece of memory of size $n$. The capability will be stored in register $r$.

@@

\subsection{Lemmas}

\subsubsection{Anti-reduction for the observation relation}

@@

@@

\begin{lemma}[Register relation downwards closed]
  \label{lem:stdrr-dc}
  \[
    n' \leq n \land \npair{w} \in \stdrr(W) \Rightarrow \npair[n']{w} \in \stdrr(W)
  \]
\end{lemma}
\begin{proof}
  By definition of $\stdrr(W)$ using Lemma~\ref{stdvr-dc}.
\end{proof}

>>>>

% Conv defs
\newcommand{\lookingat}[3]{\ensuremath{#1} \text{ is looking at } \ensuremath{#2} \text{ followed by } \ensuremath{#3}}
\newcommand{\pointstostack}[3]{\ensuremath{#1} \text{ points to stack with } \ensuremath{#2} \text{ used and } \ensuremath{#3} \text{ unused}}
\newcommand{\linksto}[4]{\ensuremath{#1} \text{ links } \ensuremath{#2} \text{ as } \ensuremath{#3} \text{ to } \ensuremath{#4}}
\newcommand{\nonlocal}[1]{\ensuremath{#1} \text{ is non-local}}

@@

              In order to compute offsets and the like, the macros need registers to keep temporary computations in. We assume such a small set of registers $\RegName_t \subseteq \RegName$ is available and that $\RegName_t$ does not contain registers explicitely named in a program nor $r_0$, $r_\stk,$ or $\pcreg$ (but clearing all registers still clears the temporary registers).

@@

              \subsection{Macro definitions}
              \begin{description}
              \item[\texttt{fetch} $r$ $f$] load the entry of the linking table corresponding to $f$ to register $r$.
              \item[\texttt{call} $r(\bar{r}_{\var{args}},\bar{r}_{\var{priv}})$] \forcenewline
                $\bar{r}_{\var{args}}$ and $\bar{r}_{\var{priv}}$ are lists of registers. An overview of this call:
                \begin{itemize}
                \item Set up activation record
                \item Create local enter capability for activation (protected return pointer)
                \item Clear unused registers
                \item Jump
                \item Upon return: Run activation code
                \end{itemize}
                A more detailed description of each of the above steps:
                \begin{description}
                \item [Set up activation record]\forcenewline
                  \begin{itemize}
                  \item Run malloc to get a piece of memory with space for:
                    \begin{itemize}
                    \item Words in $\bar{r}_{\var{priv}}$
                    \item Code return capability (opc)
                    \item Activation code
                    \end{itemize}
                  \item Store the words in $\bar{r}_{\var{priv}}$ to the activation record. 
                  \item Adjust a copy of the current pc to point to the return address in code and save it to the activation record.
                  \item Write the activation code to the activation record.
                  \end{itemize}
                \item [Create local enter capability for activation] Adjust the capability for the activation record to point to the beginning of the activation record and restrict it to a local enter-capability. Place this capability in $r_0$.
                \item [Clear unused registers]
                  Clear all the register that are not $\pcreg$, $r$, $r_0$ or in $\bar{r}_{\var{args}}$.
                \item [Jump] Jump to register $r$
                \item [Activation code] The activation code does the following:
                  \begin{itemize}
                  \item Move the stored ``private'' words in to their respective $\bar{r}_{\var{priv}}$ registers.
                    % \item Clear all registers but $r_1$, $\pcreg$, and $\bar{r}_{\var{priv}}$.
                  \item Load the return capability to $\pcreg$
                  \end{itemize}
                \end{description}
              \item[\texttt{malloc $r$ $n$}] Calls malloc to allocates a piece of memory of size $n$. The capability will be stored in register $r$. 
% Suggestion for malloc impl (r_t1 and r_t2 are registers in RegName_t), r_1 should not contain important stuff
% move r_1 n
% move r pc
% getb r_t1 r
% geta r_t2 r
% minus r_t1 r_t2 r_t1
% lea r r_t1
% move r_0 r_t1
% lea r_0 3 // 3 is offset to just after jmp
% restrict r_0 encodePerm(e)
% jmp r
% move r r_1
% move r_1 0
% move r_0 r_t1
% move r_t1 r_0

@@

\subsection{Lemmas}
\subsubsection{Anti-reduction for the observation relation}

@@

\subsubsection{Malloc macro-instruction correctness}
\begin{definition}
  We say that ``\linksto{(\reg,\ms)}{\var{key}}{j}{c_\malloc}'' 
  iff
  \begin{itemize}
  \item $\reg(pc) = \stdcap$
  \item $\ms(\start) = ((\_,\_),\start_\link,\_,\_)$
  \item $\ms(\start_\link+j) = c$
  \end{itemize}
\end{definition}

\begin{lemma}[$\mathtt{malloc}$ works]
  \label{lem:malloc-works}
  If
  \begin{itemize}
  \item \lookingat{(\reg,\ms)}{\mathtt{malloc}\;r\;k}{c_{\var{next}}}
  \item $k \geq 0$
  \item \linksto{(\reg,\ms)}{\malloc}{k}{c_\malloc}
  \item $c_\malloc$ satisfies the $\malloc$ specification with $\iota_{\malloc,0}$
  \item $W \futurestr [i \mapsto\iota_{\malloc,0}]$
  \item $\memSat{\ms}{W}$
  \item $\ms = \ms' \uplus \ms_{\var{footprint}}$
  \item $\memSat{\ms_{\var{footprint}}}{[i \mapsto W(i)]}$
  \item[Hyp-Cont]
    If
    \begin{itemize}
    \item $n' \leq n-1$
    \item $\iota_\malloc \futurewk W(i)$
    \item $\memSat[n']{\ms_{\var{footprint}}' \uplus \ms'}{W[i \mapsto \iota_\malloc]}$
    \item $\memSat[n']{\ms_{\var{footprint}}'}{[i \mapsto \iota_\malloc]}$
      \begin{equation*}
        \reg'(r') = 
        \begin{cases}
          c_{\var{next}} & r' = \pcreg \\
          ((\rwx,\glob),\start,\addrend,\addr) & r' = r \\
          \reg(r) & r' \not\in \RegName_t \union \{\pcreg, r, r_1\}
        \end{cases}
      \end{equation*}
    \item $\addrend - \start = k - 1$
    \item $\dom(\ms_{\var{alloc}}) = [\start,\addrend]$
    \item $\forall \addr \in [\start,\addrend]\ldotp \ms_{\var{alloc}}(\addr) = 0$
    \end{itemize}
    Then we have $\npair[n']{(\reg',\ms' \uplus \ms_{\var{footprint}}'  \uplus \ms_{\var{alloc}})} \in \observations(W[\iota_\malloc])$
  \end{itemize}
  Then
  \[
    \npair{(\reg,\ms)} \in \observations(W)
  \]
\end{lemma}

\subsubsection{Create closure macro-instruction correctness}
\begin{lemma}[$\mathtt{crtcls}$ works]
  If
  \begin{itemize}
  \item \lookingat{(\reg,\ms)}{\mathtt{crtcls}\;\overline{(x,r)}\;r}{c_{\var{next}}}
  \item \linksto{(\reg,\ms)}{\malloc}{k}{c_\malloc}
  \item $c_\malloc$ satisfies the $\malloc$ specification with $\iota_{\malloc,0}$
  \item $W \futurestr [i \mapsto\iota_{\malloc,0}]$
  \item $\memSat{\ms}{W}$
  \item $\ms = \ms' \uplus \ms_{\var{footprint}}$
  \item $\memSat{\ms_{\var{footprint}}}{[i \mapsto W(i)]}$
  \item[Hyp-Cont]
    If
    \begin{itemize}
    \item $n' \leq n$
    \item $\iota_\malloc \futurewk W(i)$
    \item $\memSat[n']{\ms' \uplus \ms_{\var{footprint}}'}{W[i \mapsto \iota_\malloc]}$
    \item $\memSat{\ms_{\var{footprint}}'}{[i \mapsto \iota_\malloc]}$
    \item
      \begin{equation*}
        \reg'(r') =
        \begin{cases}
          c_{\var{next}} & r' = \pcreg \\
          c_\cls = ((\entry,\glob),\start,\addrend,\start+2) & r' = r_1 \\
          \reg(r)      & r' \not\in \{\pcreg, r_1\} \union \RegName_t
        \end{cases}
      \end{equation*}
    \item $\ms_\cls = \ms_\act \uplus \ms_\env$
    \item $c_\env = ((\readwrite,\glob),\start_\env,\addrend_\env,\start_\env)$
    \item $\dom(\ms_\env) = [\start_\env,\addrend_\env]$
    \item $\ms_\env(\start_\env,\dots,\addrend_\env) = \reg(\overline{r})$
    \item Hyp-act\\ If
      \begin{itemize}
      \item $\reg''(\pcreg) = \updatePcPerm{c_\cls}$
      \end{itemize}
      Then  $\exists k \ldotp \forall \ms_f \ldotp (\reg'',\ms'' \uplus \ms_\cls \uplus \ms_f) \step[k] (\reg''',\ms'' \uplus \ms_\cls \uplus \ms_f)$
      where
        \begin{equation*}
          \reg'''(r') =
          \begin{cases}
            c_\env & r' = c_\env \\
            \updatePcPerm{\reg(r)} & r' = \pcreg \\
            \reg''(r') & r' \not\in \RegName_t
          \end{cases}
        \end{equation*}
    \end{itemize}
    Then we have $\npair[n']{(\reg',\ms' \uplus \ms_{\var{footprint}} \uplus \ms_\cls)} \in \observations(W[i \mapsto \iota_\malloc])$
  \end{itemize}
  Then 
  \[
    \npair{(\reg,\ms)} \in \observations(W)
  \]
\end{lemma}

@@

\begin{lemma}[Register relation downwards closed]
  \label{lem:stdrr-dc}
  \[
    n' \leq n \land \npair{w} \in \stdrr(W) \Rightarrow \npair[n']{w} \in \stdrr(W)
  \]
\end{lemma}
\begin{proof}
  By definition of $\stdrr(W)$ using Lemma~\ref{lem:stdvr-dc}.
\end{proof}
